---
title: "Brison, Hood & Beebower Arithmetic Attribution"
author: "Anand Krishnakumar"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brison, Hood & Beebower Arithmetic Attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PortfolioAttribution)
```

This vignette shows how to perform a single period arithmetic attribution analysis using the Brinson, Hood & Beebower model using the `PortfolioAttribution` package. For a brief introduction to portfolio performance attribution, this model, the package and references, please refer to the [introduction vignette](introduction.html).



### Define your data inputs

Say you have a simple three category portfolio & benchmark as defined below.

```{r, echo=FALSE, results='asis'}
data("sample_data", package = "PortfolioAttribution")
knitr::kable(simple_portf_1)
```

This sample data from the package is available in the package as a dataset to use.
```{r}
library(PortfolioAttribution)
data("sample_data", package = "PortfolioAttribution")
simple_portf_1
```

The portfolio & benchmark weights and returns are first extracted.

```{r}
wp = as.numeric(simple_portf_1[1,])
Rp = data.frame(simple_portf_1[2,], row.names = Sys.Date(), check.names = FALSE)
wb = as.numeric(simple_portf_1[3,])
Rb = data.frame(simple_portf_1[4,], row.names = Sys.Date(), check.names = FALSE)
```

To run attribution on this data using the Brinson, Hood & Beebower model, the following command is used.

```{r}
attribution_results = Attribution(Rp, wp, Rb, wb, bf = FALSE)
```

The different attribution effects by category as well as their totals can be inspected.

```{r}
attribution_results$Allocation
attribution_results$Selection
attribution_results$Interaction
```

As expected the total arithmetic excess return equals the sum of the total for each attribution effect

```{r}
excess_returns = attribution_results$`Excess returns`
excess_returns

total_effects = attribution_results$Allocation[1,"Total"] + attribution_results$Selection[1,"Total"] + attribution_results$Interaction[1,"Total"]
total_effects
```


