---
title: "Contribution Analysis"
author: "[Anand Krishnakumar](https://github.com/anandkkumar)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contribution Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PortfolioAttribution)
```

Contribution analysis is a concept that is related to attribution analysis. This vignette shows how to retrieve contribution data for the portfolio and benchmark under review with the `PortfolioAttribution` package.

### Define the data inputs

```{r, echo=FALSE}
data("sample_data", package = "PortfolioAttribution")
```

Say we have a relatively large portfolio & benchmark data set at the security level. This example data set in the package has `r NCOL(multi_period_portf_1$Rp)` assets over `r NROW(multi_period_portf_1$Rp)` monthly periods. The column names are security identifiers and the row indexes are month-end dates.

The first few periods & assets are shown below.

```{r}
library(PortfolioAttribution)
data("sample_data", package = "PortfolioAttribution")
# Portfolio weights
multi_period_portf_1$wp[1:3,1:4]

# Portfolio returns
multi_period_portf_1$Rp[1:3,1:4]

# Benchmark weights
multi_period_portf_1$wb[1:3,1:4]

# Benchmark returns
multi_period_portf_1$Rb[1:3,1:4]
```

#### Contribution analysis

To include contribution data in the results when performing attribution analysis, the `contribution` parameter is set to `TRUE` (the default is `FALSE`).

```{r}
attribution_results = Attribution(multi_period_portf_1$Rp, multi_period_portf_1$wp, 
                                  multi_period_portf_1$Rb, multi_period_portf_1$wb, 
                                  contribution = TRUE)
```

In addition, to reporting the attribution effects, the result also contains the contribution for each period by category for both the portfolio and the benchmark.

```{r}
names(attribution_results)
```

The contribution data for the first few periods & assets are shown below.

```{r}
# Portfolio contribution to return
attribution_results$`Portfolio contribution to return`[1:3,1:4]

# Benchmark contribution to return
attribution_results$`Benchmark contribution to return`[1:3,1:4]
```
