---
title: "Multi-period geometric attribution"
author: "[Anand Krishnakumar](https://github.com/anandkkumar)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-period geometric attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PortfolioAttribution)
```

This vignette shows how to perform a multi-period geometric attribution analysis on a portfolio using the the `PortfolioAttribution` package. Unlike in the case of geometric attribution, geometric attribution effects naturally compound over time and there is no need to smooth the effects in order for the the total attribution effects to match the total geometric excess returns over all periods.

### <ins>Define the data inputs</ins>

```{r, echo=FALSE}
data("sample_data", package = "PortfolioAttribution")
```

Say we have a relatively large portfolio & benchmark data set at the security level. This example data set in the package has `r NCOL(multi_period_portf_1$Rp)` assets over `r NROW(multi_period_portf_1$Rp)` monthly periods. The column names are security identifiers and the row indexes are month-end dates. As outlined in the [introduction vignette](introduction.html), the attribution package requires that the number of assets be the same for all periods and between the portfolio & benchmark.

The first few periods & assets are shown below.

```{r}
library(PortfolioAttribution)
data("sample_data", package = "PortfolioAttribution")
# Portfolio weights
multi_period_portf_1$wp[1:3,1:4]

# Portfolio returns
multi_period_portf_1$Rp[1:3,1:4]

# Benchmark weights
multi_period_portf_1$wb[1:3,1:4]

# Benchmark returns
multi_period_portf_1$Rb[1:3,1:4]
```

Since the package uses an end-of-period weight convention, the index of the weight data is the month-end prior to that of the return data.

```{r}
index(multi_period_portf_1$wp[1,])
index(multi_period_portf_1$Rp[1,])

index(multi_period_portf_1$wb[1,])
index(multi_period_portf_1$Rb[1,])
```


### <ins>Multi-period attribution</ins>

The following command is executed to perform a multi-period geometric attribution analysis.

```{r}
attribution_results = Attribution(multi_period_portf_1$Rp, multi_period_portf_1$wp, 
                                  multi_period_portf_1$Rb, multi_period_portf_1$wb, 
                                  geometric = TRUE) 
```

The different attribution effects by category as well as their totals can be inspected. The first few periods and effects for some of the assets and their totals are shown below.

```{r}
attribution_results$Allocation[1:3, c(1:3, NCOL(attribution_results$Allocation))]
attribution_results$Selection[1:3, c(1:3, NCOL(attribution_results$Allocation))]
```

The data set also reports the total portfolio & benchmark returns for each period and so the total geometric excess return can be computed from them.

```{r}
total_portfolio_return = prod(1 + multi_period_portf_1$periodicPortReturns) - 1
total_portfolio_return

total_benchmark_return = prod(1 + multi_period_portf_1$periodicBmkReturns) - 1
total_benchmark_return

total_geometric_excess_return = (1 + total_portfolio_return)/(1 + total_benchmark_return) - 1
total_geometric_excess_return

```

With geometric attribution, the total allocation & selection effects across all periods <ins>compound</ins> to give the total effect, which as expected equals the total geometric excess return.

```{r}
total_effect = (1 + attribution_results$Allocation["Total", "Total"]) * 
               (1 + attribution_results$Selection["Total", "Total"]) - 1
all.equal(total_geometric_excess_return, total_effect)
```
