---
title: "Run-times for attribution analysis"
author: "[Anand Krishnakumar](https://github.com/anandkkumar)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Run-times for attribution analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PortfolioAttribution)
```

This vignette demonstrates how to measure the computation times on a large data set of portfolio & benchmark weights and returns. It is intended to be illustrative and does not represent a rigorous approach to benchmarking the performance of the functions. The `system.time` command is employed and the results will naturally vary based on your computing platform.

### <ins>Define the data inputs</ins>

```{r, echo=FALSE}
data("sample_data", package = "PortfolioAttribution")
```

Say we have two relatively large portfolio & benchmark data sets at the security level. This first example in the data set in the package has `r NCOL(multi_period_portf_1$Rp)` assets over `r NROW(multi_period_portf_1$Rp)` monthly periods and the second example in the data set has `r NCOL(multi_period_portf_3$Rp)` assets over `r NROW(multi_period_portf_3$Rp)` monthly periods. The column names are security identifiers and the row indexes are month-end dates.

The first few periods & assets are shown below.

```{r}
library(PortfolioAttribution)
data("sample_data", package = "PortfolioAttribution")
# Portfolio 1 weights
multi_period_portf_1$wp[1:3,1:5]

# Portfolio 1 returns
multi_period_portf_1$Rp[1:3,1:5]

# Benchmark 1 weights
multi_period_portf_1$wb[1:3,1:5]

# Benchmark 1 returns
multi_period_portf_1$Rb[1:3,1:5]

# Portfolio 2 weights
multi_period_portf_3$wp[1:3,1:5]

# Portfolio 2 returns
multi_period_portf_3$Rp[1:3,1:5]

# Benchmark 2 weights
multi_period_portf_3$wb[1:3,1:5]

# Benchmark 2 returns
multi_period_portf_3$Rb[1:3,1:5]
```

### <ins>Run-time analysis</ins>

The following command is executed to run & time a multi-period arithmetic attribution analysis using the default _Brinson and Fachler [-@brinson1985measuring]_ model and adopting a top-down approach with the default smoothing algorithm of _GRAP [-@grap1997]_ on the first example data set (please refer to the [multi-period arithmetic attribution vignette](multi-period-arithmetic.html) for more details on smoothing).

```{r}
system.time(Attribution(multi_period_portf_1$Rp, multi_period_portf_1$wp, 
                        multi_period_portf_1$Rb, multi_period_portf_1$wb, 
                        method = "top.down"))
```


The following command is executed to run & time a multi-period geometric attribution analysis.

```{r}
system.time(Attribution(multi_period_portf_1$Rp, multi_period_portf_1$wp, 
                        multi_period_portf_1$Rb, multi_period_portf_1$wb, 
                        geometric = TRUE))
```

The following command is executed to run & time a multi-period arithmetic attribution analysis using the default _Brinson and Fachler [-@brinson1985measuring]_ model and adopting a top-down approach with the default smoothing algorithm of _GRAP [-@grap1997]_ on the second example data set.

```{r}
system.time(Attribution(multi_period_portf_3$Rp, multi_period_portf_3$wp, 
                        multi_period_portf_3$Rb, multi_period_portf_3$wb, 
                        method = "top.down"))
```

The following command is executed to run & time a multi-period geometric attribution analysis.

```{r}
system.time(Attribution(multi_period_portf_3$Rp, multi_period_portf_3$wp, 
                        multi_period_portf_3$Rb, multi_period_portf_3$wb, 
                        geometric = TRUE))
```

In general,

* the arithmetic attribution analysis takes longer than the geometric attribution analysis largely due to the need for computing smoothed effects for _each_ period to arrive at the total effects for arithmetic attribution
* the longer the analysis period, the longer it takes to do the attribution analysis

### <ins>References</ins>
