---
title: "Single period geometric attribution"
author: "[Anand Krishnakumar](https://github.com/anandkkumar)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single period geometric attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PortfolioAttribution)
```

This vignette shows how to perform a single period geometric attribution analysis using the `PortfolioAttribution` package and illustrates the point about the weight inputs being treated as __end-of-period__ weights. For a brief introduction to portfolio performance attribution, this model, the package and references, please refer to the [introduction vignette](introduction.html).



### Define the data inputs

```{r, echo=FALSE}
data("sample_data", package = "PortfolioAttribution")
```

Say we have a relatively large portfolio & benchmark data set at the security level. This example data set in the package has `r NCOL(multi_period_portf_1$Rp)` assets over `r NROW(multi_period_portf_1$Rp)` monthly periods. The column names are security identifiers and the row indexes are month-end dates. As outlined in the [introduction vignette](introduction.html), the attribution package requires that the number of assets be the same for all periods and between the portfolio & benchmark.

The first few periods & assets are shown below.

```{r}
data("sample_data", package = "PortfolioAttribution")
# Portfolio weights
multi_period_portf_1$wp[1:3,1:10]

# Portfolio returns
multi_period_portf_1$Rp[1:3,1:10]

# Benchmark weights
multi_period_portf_1$wb[1:3,1:10]

# Benchmark returns
multi_period_portf_1$Rb[1:3,1:10]
```

To perform a single period attribution analysis just the first observation for the portfolio & benchmark is extracted. Since the package uses an end-of-period weight convention, the index of the weight data is the month-end prior to that of the return data.

```{r}
wp = multi_period_portf_1$wp[1,]
Rp = multi_period_portf_1$Rp[1,]
index(wp)
index(Rp)

wb = multi_period_portf_1$wb[1,]
Rb = multi_period_portf_1$Rb[1,]
index(wb)
index(Rb)
```

#### Geometric attribution

To run geometric attribution analysis on this data the following command is used.

```{r}
attribution_results = Attribution(Rp, wp, Rb, wb, geometric = TRUE)
```

The total allocation & selection effects for all securities is available in the _Total_ column for the period. In this case, since we had no categories and all the data was at the security level, the selection effects are by construction all zero.

```{r}
attribution_results$Allocation[1, "Total"]
attribution_results$Selection[1, "Total"]
```

The data set also reports the total portfolio & benchmark return for each period and so the geometric excess return can be computed from that

```{r}
geometric_excess_return = as.numeric((1+multi_period_portf_1$periodicPortReturns[1,])/
                                       (1+multi_period_portf_1$periodicBmkReturns[1,]) - 1)
geometric_excess_return
```

With geometric attribution, the total allocation & selection effects <ins>compound</ins> to give the total effect, which as expected equals the total geometric excess return.

```{r}
total_effect = (1+attribution_results$Allocation[1, "Total"])*(1+attribution_results$Selection[1, "Total"])-1
all.equal(geometric_excess_return, total_effect)
```
