---
title: "Single period arithmetic attribution with currency effects"
author: "[Anand Krishnakumar](https://github.com/anandkkumar)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: introduction.bib
vignette: >
  %\VignetteIndexEntry{Single period arithmetic attribution with currency effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style type="text/css">
body {
  max-width: 100% !important;
}
table {
  margin-left: 0 !important;
}
</style>

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(magrittr)
library(PortfolioAttribution)
```

This vignette shows how to perform a single period arithmetic attribution analysis on a multi-currency portfolio using the the `PortfolioAttribution` package with two different approaches. For more methodological details please refer to the [introduction vignette](introduction.html) and the [multi-currency introduction vignette](introduction-multi-currency.html).


### <ins>Simple differences</ins>

Some practitioners & attribution platforms simply consider the _currency effect_ to be the difference in effects between running the attribution analysis with returns in the base currency and again with returns in the local currency.

#### Define the data inputs

Consider a portfolio & benchmark data categorized into sectors as shown below.

```{r, echo=FALSE, results='asis'}
data("sample_data", package = "PortfolioAttribution")
knitr::kable(simple_multi_curr_portf_1)
```

This sample data from the package is available in the package as a data set to use.
```{r}
data("sample_data", package = "PortfolioAttribution")
simple_multi_curr_portf_1
```

The weights, local and base currency returns for portfolio & benchmark are first extracted.

```{r}
wp = as.numeric(simple_multi_curr_portf_1[1,])
Rpl = data.frame(simple_multi_curr_portf_1[2,], row.names = Sys.Date(), check.names = FALSE)
Rp = data.frame(simple_multi_curr_portf_1[3,], row.names = Sys.Date(), check.names = FALSE)
wb = as.numeric(simple_multi_curr_portf_1[4,])
Rbl = data.frame(simple_multi_curr_portf_1[5,], row.names = Sys.Date(), check.names = FALSE)
Rb = data.frame(simple_multi_curr_portf_1[6,], row.names = Sys.Date(), check.names = FALSE)
```

#### Brison and Fachler top-down attribution

The following commands are executed to perform attribution analysis on this data using the Brinson and Fachler model and adopting a top-down approach (where the interaction effect is combined with the selection effect) successively using local currency returns and base currency returns.

```{r}
attribution_results_local = Attribution(Rpl, wp, Rbl, wb, method = "top.down")
attribution_results = Attribution(Rp, wp, Rb, wb, method = "top.down")
```

As expected, for both local and base currencies the total arithmetic excess return equals the sum of the total attribution effects

```{r}
excess_returns_local = attribution_results_local$`Excess returns`
excess_returns_local

total_effects_local = attribution_results_local$Allocation[1,"Total"] + attribution_results_local$Selection[1,"Total"]
total_effects_local

excess_returns = attribution_results$`Excess returns`
excess_returns

total_effects = attribution_results$Allocation[1,"Total"] + attribution_results$Selection[1,"Total"]
total_effects
```

The currency effects for each category & the total currency effect are computed as the difference between the total attribution effects using base currency returns and those using local currency returns.

```{r}
currency_effects = attribution_results$Allocation + attribution_results$Selection  - attribution_results_local$Allocation - attribution_results_local$Selection
```

Finally, the total arithmetic excess returns in the base currency is equal to the sum of the total allocation & selection effects for all categories using the local currency returns and the total currency effect.

```{r}
all.equal(as.numeric(attribution_results_local$Allocation[1,"Total"] + 
                     attribution_results_local$Selection[1,"Total"] + 
                     currency_effects[1,"Total"]), 
          as.numeric(excess_returns))
```


Shown below is a summary of the all the effects including currency effects

```{r echo=FALSE, results='asis'}
results_summary = as.data.frame(rbind(attribution_results$Allocation, 
                                      attribution_results$Selection, 
                                      attribution_results_local$Allocation, 
                                      attribution_results_local$Selection, 
                                      currency_effects))
rownames(results_summary) = c("Base Currency Allocation", "Base Currency Selection",
                              "Local Currency Allocation", "Local Currency Selection", 
                              "Curreny effect")
results_summary %>% kableExtra::kable("html") %>% 
  kableExtra::row_spec(3:5, background = "#F7F7F7") %>% 
  kableExtra::column_spec(13, bold = TRUE)
```


### <ins>Ankrim and Hensel method</ins>

The package also supports computing currency effects using the methodology described by _Ankrim and Hensel [-@ankrim1994multicurrency]_.

#### Define the data inputs

Consider a portfolio & benchmark data categorized into three regions as shown below.

```{r, echo=FALSE, results='asis'}
data("sample_data", package = "PortfolioAttribution")
knitr::kable(simple_multi_curr_portf_2)
knitr::kable(simple_multi_curr_portf_2_spot_forward_rates)
```

This sample data from the package is available in the package as a data set to use.
```{r}
data("sample_data", package = "PortfolioAttribution")
simple_multi_curr_portf_2

simple_multi_curr_portf_2_spot_forward_rates
```

The weights returns for portfolio & benchmark categories, the weights for the portfolio & benchmark forward contracts and the spot & forward rates for the relevant current pairs are extracted from the sample data sets.

```{r}
wp = as.numeric(simple_multi_curr_portf_2[1,])
Rp = data.frame(simple_multi_curr_portf_2[2,], row.names = Sys.Date(), check.names = FALSE)
wb = as.numeric(simple_multi_curr_portf_2[3,])
Rb = data.frame(simple_multi_curr_portf_2[4,], row.names = Sys.Date(), check.names = FALSE)
wpf = as.numeric(simple_multi_curr_portf_2[5,])
wbf = as.numeric(simple_multi_curr_portf_2[6,])
S = data.frame(simple_multi_curr_portf_2_spot_forward_rates[1:2,], row.names = c(Sys.Date(), Sys.Date()+1), check.names = FALSE)
Fp = data.frame(simple_multi_curr_portf_2_spot_forward_rates[3:4,], row.names = c(Sys.Date(), Sys.Date()+1), check.names = FALSE)
Fb = data.frame(simple_multi_curr_portf_2_spot_forward_rates[5:6,], row.names = c(Sys.Date(), Sys.Date()+1), check.names = FALSE)
```

It is important to note that the spot and forward rates data inputs need to have one more period than that of the portfolio & benchmark return data set. In addition, the first date of the data set should match up with that of the portfolio returns.

```{r}
Rp;Rb

S;Fp;Fb
```

#### Currency attribution

The following command is executed to perform standard attribution analysis with currency effects on this data set using the Ankrim and Hensel methodology.

```{r}
attribution_results = Attribution(Rp, wp, Rb, wb, wpf, wbf, S, Fp, Fb)
```

The two currency effects reported are _Currency Management_ and _Forward Premium_ in addition to the usual attribution effects of allocation, selection and interaction

```{r}
names(attribution_results)
attribution_results$`Currency management`
attribution_results$`Forward Premium`
```

As expected the total arithmetic excess return can be decomposed into the sum of the three standard attribution effects and the two currency effects.

```{r}
attribution_results$`Excess returns`

all.equal(as.numeric(attribution_results$Allocation[,"Total"] + 
                     attribution_results$Selection[,"Total"] + 
                     attribution_results$Interaction[,"Total"] + 
                     attribution_results$`Currency management`[,"Total"] +
                     attribution_results$`Forward Premium`[,"Total"]),
          as.numeric(attribution_results$`Excess returns`))
```

### <ins>References</ins>
